---
title: "Data Engineering and Visualisation Assessment 3"
author:
   - name: Sreekantom Sai Saketh (2503848)
   - name: Tng Qi Long Felix (2500906)
   - name: Aruna Pressanna Dhevakash (2501146)
   - name: Tay Rui Feng (2500458)
   - name: Ang Jia Jun Ryan (2503294)
date: 15 July 2026
format:
  html:
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 2

execute:
  echo: true
  warning: false
  message: false
---

# Library Setup
```{r}
library(tidyverse)
library(gt)
library(readxl)
library(tidyquant)
```

# Task 1: English Premier League Table 2019–2020

## Introduction

This report analyses the English Premier League match results using R, the **tidyverse** suite and the **gt** package. The first part generates the final league table based on total points, goal difference and goals scored. The second part proposes a **Home Advantage Index (HAI)** to quantify how much each team benefits from playing at home.

The analysis uses match-level results from all 20 Premier League teams and presents the findings through ranked and professionally formatted tables.

```{r}
results <- read_csv("epl_results.csv", show_col_types = FALSE)
teams <- read_csv("epl_teams.csv", show_col_types = FALSE)
```

#  EPL Match Analysis

## Generating the League Table

### Transforming the Match Results

The original dataset stores match results in a wide format, where each away team is represented as a separate column. To simplify the analysis, the data is reshaped into a long format so that each row represents a single match between a home team and an away team.

```{r}
matches <- results |>
  pivot_longer(
    cols = -home_team,
    names_to = "away_team",
    values_to = "result"
  ) |>
  filter(result != "-")

matches
```

## Separating Home and Away Goals

Match results are initially stored as text in the format **home goals–away goals** (e.g., "3-2"). This step separates the score into two numerical columns, allowing the goals scored by each team to be analysed and compared.

```{r}
matches <- matches |>
  separate_wider_delim(
    result,
    delim = "-",
    names = c("home_goals", "away_goals")
  ) |>
  mutate(
    home_goals = as.integer(home_goals),
    away_goals = as.integer(away_goals)
  )

matches
```

## Calculating Team Results

Each match is converted into two team-level records: one for the home team and one for the away team. Points are then awarded according to Premier League rules, where a win earns three points, a draw earns one point and a loss earns zero points.

```{r}
home_records <- matches |>
  transmute(
    team = home_team,
    goals_scored = home_goals,
    goals_conceded = away_goals,
    points = case_when(
      home_goals > away_goals ~ 3,
      home_goals == away_goals ~ 1,
      TRUE ~ 0
    )
  )

away_records <- matches |>
  transmute(
    team = away_team,
    goals_scored = away_goals,
    goals_conceded = home_goals,
    points = case_when(
      away_goals > home_goals ~ 3,
      away_goals == home_goals ~ 1,
      TRUE ~ 0
    )
  )
```

```{r}
team_records <- bind_rows(home_records, away_records)

team_records
```

## Summarizing the League Results

The individual match records are aggregated to calculate each team's season statistics, including total points, goals scored, goals conceded and goal difference. These values are then used to rank the teams according to the official Premier League standings.

```{r}
league_table <- team_records |>
  mutate(
    win = if_else(points == 3, 1, 0),
    draw = if_else(points == 1, 1, 0),
    loss = if_else(points == 0, 1, 0)
  ) |>
  group_by(team) |>
  summarise(
    MP = n(),
    W = sum(win),
    D = sum(draw),
    L = sum(loss),
    GF = sum(goals_scored),
    GA = sum(goals_conceded),
    GD = GF - GA,
    Pts = sum(points),
    .groups = "drop"
  ) |>
  left_join(
    teams,
    by = c("team" = "abbr")
  ) |>
  select(
    Team = name,
    MP,
    W,
    D,
    L,
    GF,
    GA,
    GD,
    Pts
  ) |>
  arrange(
    desc(Pts),
    desc(GD),
    desc(GF)
  ) |>
  mutate(
    Pos = row_number()
  ) |>
  select(
    Pos,
    Team,
    MP,
    W,
    D,
    L,
    GF,
    GA,
    GD,
    Pts
  )

league_table
```

## Final League Table

The final league table presents the ranked standings of all teams using a professionally formatted table. The top four teams are highlighted as Champions League qualifiers, while the bottom three teams are highlighted as the relegation zone.

```{r}
league_table |>
  gt() |>
  tab_header(
    title = md("**English Premier League Final Table**"),
    subtitle = "Ranked by points, then goal difference, then goals scored"
  ) |>
  cols_label(
    Pos = "Pos",
    Team = "Team",
    MP = "MP",
    W = "W",
    D = "D",
    L = "L",
    GF = "GF",
    GA = "GA",
    GD = "GD",
    Pts = "Pts"
  ) |>
  cols_align(
    align = "left",
    columns = Team
  ) |>
  cols_align(
    align = "center",
    columns = c(Pos, MP, W, D, L, GF, GA, GD, Pts)
  ) |>
  fmt_number(
    columns = c(Pos, MP, W, D, L, GF, GA, GD, Pts),
    decimals = 0
  ) |>
  fmt(
    columns = GD,
    fns = function(x) {
      ifelse(x > 0, paste0("+", x), x)
    }
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#1F4E78"),
      cell_text(
        color = "white",
        weight = "bold"
      )
    ),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#D9EAF7"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Pos <= 4
    )
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#FCE8E6"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Pos >= 18
    )
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = Pts
    )
  ) |>
  tab_style(
    style = cell_borders(
      sides = "bottom",
      color = "#1F4E78",
      weight = px(2)
    ),
    locations = cells_body(
      rows = Pos == 4
    )
  ) |>
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "#B22222",
      weight = px(2)
    ),
    locations = cells_body(
      rows = Pos == 18
    )
  ) |>
  opt_row_striping() |>
  opt_table_outline() |>
  tab_options(
    table.font.size = px(14),
    heading.title.font.size = px(24),
    heading.subtitle.font.size = px(14),
    heading.align = "center",
    column_labels.font.size = px(13),
    row.striping.background_color = "#F7F7F7",
    data_row.padding = px(7),
    table.width = pct(100)
  ) |>
  tab_source_note(
    source_note = md(
      "**MP:** Matches Played · **W:** Wins · **D:** Draws · **L:** Losses ·  
      **GF:** Goals Scored · **GA:** Goals Against · **GD:** Goal Difference · **Pts:** Points"
    )
  ) |>
  tab_source_note(
    source_note = md(
      "**Blue:** Champions League qualification · **Red:** Relegation zone"
    )
  )
```

# Designing a Home Advantage Index

## Metric Definition

The Home Advantage Index measures how much better a team performs at home compared with away. It is calculated by subtracting away points per game from home points per game. A higher positive value indicates a stronger home advantage, while a value close to zero suggests similar performance at home and away. A negative value means the team earns more points per game away from home.

```{r}
home_performance <- matches |>
  transmute(
    team = home_team,
    venue = "Home",
    goals_scored = home_goals,
    goals_conceded = away_goals,
    points = case_when(
      home_goals > away_goals ~ 3,
      home_goals == away_goals ~ 1,
      TRUE ~ 0
    )
  )

away_performance <- matches |>
  transmute(
    team = away_team,
    venue = "Away",
    goals_scored = away_goals,
    goals_conceded = home_goals,
    points = case_when(
      away_goals > home_goals ~ 3,
      away_goals == home_goals ~ 1,
      TRUE ~ 0
    )
  )

venue_performance <- bind_rows(
  home_performance,
  away_performance
)
```

## Calculating the Home Advantage Index

> To calculate the Home Advantage Index (HAI), each team's home and away performances are analysed separately. The average points earned per game at home and away are calculated, and the difference between these two values is used as the HAI. A larger positive value indicates that a team performs significantly better when playing at home.

```{r}
home_advantage_table <- venue_performance |>
  group_by(team, venue) |>
  summarise(
    games_played = n(),
    total_points = sum(points),
    points_per_game = total_points / games_played,
    .groups = "drop"
  ) |>
  select(
    team,
    venue,
    points_per_game
  ) |>
  pivot_wider(
    names_from = venue,
    values_from = points_per_game
  ) |>
  mutate(
    metric_value = Home - Away
  ) |>
  left_join(
    teams,
    by = c("team" = "abbr")
  ) |>
  arrange(desc(metric_value)) |>
  mutate(
    metric_rank = row_number()
  ) |>
  transmute(
  metric_rank,
  team = name,
  metric_value = round(metric_value, 2),
  home_ppg = round(Home, 2),
  away_ppg = round(Away, 2)
  )

home_advantage_table
```

The calculated HAI values are ranked from highest to lowest and presented in a formatted table. Additional columns showing home and away points per game are included to provide context and allow the results to be interpreted more easily.

```{r}
home_advantage_table |>
  gt() |>
  tab_header(
    title = md("**Home Advantage Index Rankings**"),
    subtitle = "Higher values indicate a stronger home advantage"
  ) |>
  cols_label(
    metric_rank = "Rank",
    team = "Team",
    metric_value = "HAI",
    home_ppg = "Home PPG",
    away_ppg = "Away PPG"
  ) |>
  cols_align(
    align = "center",
    columns = c(metric_rank, metric_value, home_ppg, away_ppg)
  ) |>
  cols_align(
    align = "left",
    columns = team
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#1F4E78"),
      cell_text(color = "white", weight = "bold")
    ),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#D9EAF7"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = metric_rank <= 3
    )
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = metric_value
    )
  ) |>
  data_color(
    columns = metric_value,
    palette = c("#FCE8E6", "#FFFFFF", "#D9EAF7")
  ) |>
  opt_row_striping() |>
  tab_options(
    table.width = pct(85),
    heading.align = "center",
    table.font.size = px(15),
    data_row.padding = px(8)
  ) |>
  tab_source_note(
    source_note = md(
      "**HAI = Home Points per Game − Away Points per Game**"
    )
  )
```

The proposed Home Advantage Index provides a meaningful and interpretable measure of home advantage by comparing points earned per game at home and away. Unlike total points, points per game normalises performance across an equal number of fixtures, ensuring fair comparisons between teams. The results indicate that Tottenham Hotspur exhibits the strongest home advantage, while clubs with lower HAI values perform more consistently across venues. Using points per game rather than total points ensures the comparison remains fair and independent of league position, making the metric suitable for comparing teams with different overall strengths.

# Task 2: Investment Market Price Analysis

## 2.1 DJIA Performance

## Import and clean the data (2%)

```{r}
raw_dj <- read_excel("dja-performance-report-monthly.xlsx", skip = 5)

monthly_dj <- raw_dj |>
  filter(`Index Name` == "Dow Jones Industrial Average") |>
  transmute(
    date = as.Date(`Effective Date`, format = "%m/%d/%Y"),
    close_value = as.numeric(`Close Value`)
  )

head(monthly_dj, 10)
```

## Identify the gap months (3%)

**Historical reason:** The New York Stock Exchange closed for over
four months (31 Jul–12 Dec 1914) at the outbreak of World War I, fearing
foreign investors would dump U.S. securities and drain gold reserves. No
trading meant no DJIA prices during that window, creating the gap in the
data.

```{r}
monthly_dj_check <- monthly_dj |>
  arrange(date) |>
  mutate(
    month = month(date),
    year = year(date),
    months_to_next_row = interval(date, lead(date)) %/% months(1)
  )

# The gap(s): any row where more than 1 month passes before the next entry
gap_start <- monthly_dj_check |>
  filter(months_to_next_row > 1)

gap_start
```

## Visualizing DJIA performance by month (10%)

- `rel_change_pct`: "how much did this month's close change vs last month's,
  in percent?"
- `month_abb`: turns the numeric month (1–12) into "Jan", "Feb", ... and makes
  it an **ordered factor**.

```{r}
monthly_dj <- monthly_dj |>
  arrange(date) |>
  mutate(
    rel_change_pct = (close_value / lag(close_value) - 1) * 100,
    month_abb = factor(month.abb[month(date)], levels = month.abb, ordered = TRUE)
  )

head(monthly_dj, 10)
```

### Visualization A: violin + box plot, with a line connecting medians

```{r viz-a}
monthly_dj_clean <- monthly_dj |> drop_na(rel_change_pct)

medians <- monthly_dj_clean |>
  group_by(month_abb) |>
  summarize(median_change = median(rel_change_pct), .groups = "drop")

ggplot(monthly_dj_clean, aes(x = month_abb, y = rel_change_pct)) +
  geom_violin(fill = "#a6cee3", color = "#1f78b4", trim = TRUE) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  geom_line(data = medians, aes(x = month_abb, y = median_change, group = 1),
            color = "firebrick", linewidth = 0.8) +
  geom_point(data = medians, aes(x = month_abb, y = median_change),
             color = "firebrick", size = 1.8) +
  labs(
    title = "DJIA Monthly Relative Change by Calendar Month",
    subtitle = "Violin shows full distribution shape; box shows quartiles; red line tracks the median",
    x = "Month", y = "Relative Change (%)"
  ) +
  coord_cartesian(ylim = quantile(monthly_dj_clean$rel_change_pct, c(0.01, 0.99))) +
  theme_minimal()
```

### Visualization B: faceted frequency polygons + rug plot

`facet_wrap()` splits one big plot into a grid of small ones, one per group —
here, one panel per month.

```{r viz-b}
means <- monthly_dj_clean |>
  group_by(month_abb) |>
  summarize(mean_change = mean(rel_change_pct), .groups = "drop")

ggplot(monthly_dj_clean, aes(x = rel_change_pct)) +
  geom_freqpoly(bins = 30, color = "#1f78b4", linewidth = 0.7) +
  geom_rug(sides = "b", alpha = 0.3, color = "#1f78b4") +
  geom_vline(data = means, aes(xintercept = mean_change),
             linetype = "dashed", color = "firebrick") +
  facet_wrap(~ month_abb, nrow = 4, ncol = 3) +
  labs(
    title = "Distribution of DJIA Monthly Relative Change, by Month",
    subtitle = "Dashed red line = mean change for that month",
    x = "Relative Change (%)", y = "Count"
  ) +
  theme_minimal()
```

# 2.2 Comparing DJIA, BTC, and Gold Prices (25%)

## Import BTC and gold data (5%)

Import the BTC and gold price data since 2016 using the tq_transmute() function from the tidyquant package, and create `monthly_btc` and `monthly_gold` tibbles

```{r}
ticker_btc <- "BTC-USD"
ticker_gold <- "GC=F"
start_date <- "2016-01-01"
end_date <- "2026-06-30"

# BTC Data
monthly_btc <- tq_get(ticker_btc, from = start_date, to = end_date) |>
  drop_na(adjusted) |>
  tq_transmute(select = adjusted, mutate_fun = to.monthly, indexAt = "lastof")
# Gold Data
monthly_gold <- tq_get(ticker_gold, from = start_date, to = end_date) |>
  drop_na(adjusted) |>
  tq_transmute(select = adjusted, mutate_fun = to.monthly, indexAt = "lastof")

head(monthly_btc, 10)
head(monthly_gold, 10)
```

## Combined visualization (10%)

**Problem:** DJIA trades in tens of thousands, Gold in the
thousands, and BTC swings from under $1,000 to over $60,000 across this
period. Plot them raw on one chart and DJIA/BTC will flatten gold into an
invisible line at the bottom — like putting an elephant, a dog, and a mouse
on the same bathroom scale display.

**Solution: Indexing:** Rescale every series so it starts at 100. Now
we're not asking "what's the dollar price?" but "how much has this grown
since Jan 2016?" — a fair comparison. We also use a log y-axis, since BTC's
growth is multiplicative (10x, 100x) rather than additive, and log scales
make percentage moves visually comparable regardless of the starting size.

```{r}
dj_long <- monthly_dj |>
  filter(date >= as.Date(start_date)) |>
  transmute(month = floor_date(date, "month"), close_value, asset = "DJIA")

btc_long <- monthly_btc |>
  transmute(month = floor_date(as.Date(date), "month"), close_value = adjusted, asset = "BTC")

gold_long <- monthly_gold |>
  transmute(month = floor_date(as.Date(date), "month"), close_value = adjusted, asset = "Gold")

combined <- bind_rows(dj_long, btc_long, gold_long) |>
  drop_na(close_value)

# keep only the date range where ALL THREE assets have data
range_by_asset <- combined |>
  group_by(asset) |>
  summarize(min_month = min(month), max_month = max(month), .groups = "drop")

common_start <- max(range_by_asset$min_month)
common_end <- min(range_by_asset$max_month)

combined <- combined |>
  filter(month >= common_start, month <= common_end)

# index each series to 100 at the common start date
combined_indexed <- combined |>
  group_by(asset) |>
  arrange(month) |>
  mutate(index_100 = close_value / first(close_value) * 100) |>
  ungroup()

ggplot(combined_indexed, aes(x = month, y = index_100, color = asset)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  scale_color_manual(values = c(DJIA = "#1f78b4", BTC = "#e6550d", Gold = "#b15928")) +
  labs(
    title = "DJIA vs BTC vs Gold: Growth Since Jan 2016",
    subtitle = "Indexed to 100 at start, log scale so % moves are visually comparable",
    x = "Month", y = "Indexed Price (log scale, start = 100)", color = "Asset"
  ) +
  theme_minimal()
```

## Volatility analysis: largest monthly drawdown (10%)

```{r}
compute_drawdown <- function(prices) {
  running_max <- cummax(prices)
  drawdown_pct <- (prices - running_max) / running_max * 100
  max_drawdown <- abs(min(drawdown_pct))
  max_drawdown
}

# Sanity check: peak of 120 -> trough of 90 is the biggest fall = -25%...
# wait, actually walk through it: 100 -> 60 (peak so far 100, DD -40%)
#                                  60 -> 120 (new peak, DD 0%)
#                                  120 -> 90 (peak 120, DD -25%)
# so the worst drawdown in this vector is -40%, matching the expected answer.
prices_test <- c(100, 60, 120, 90)
compute_drawdown(prices_test)  # should print 40
```

Now we extend the function slightly so it also tells us *when* the peak and
trough happened — useful for tying the numbers back to real events.

```{r drawdown-details}
compute_drawdown_details <- function(dates, prices) {
  running_max <- cummax(prices)
  drawdown_pct <- (prices - running_max) / running_max * 100
  trough_idx <- which.min(drawdown_pct)
  peak_idx <- max(which(prices[1:trough_idx] == running_max[trough_idx]))

  tibble(
    max_drawdown_pct = abs(drawdown_pct[trough_idx]),
    peak_date = dates[peak_idx],
    peak_price = prices[peak_idx],
    trough_date = dates[trough_idx],
    trough_price = prices[trough_idx]
  )
}

dj_since_2016 <- monthly_dj |>
  filter(date >= as.Date(start_date)) |>
  arrange(date)

drawdowns_dj <- compute_drawdown_details(dj_since_2016$date, dj_since_2016$close_value)
drawdowns_gold <- compute_drawdown_details(as.Date(monthly_gold$date), monthly_gold$adjusted)
drawdowns_btc <- compute_drawdown_details(as.Date(monthly_btc$date), monthly_btc$adjusted)

biggest_drawdown_dj <- drawdowns_dj$max_drawdown_pct
biggest_drawdown_gold <- drawdowns_gold$max_drawdown_pct
biggest_drawdown_btc <- drawdowns_btc$max_drawdown_pct

biggest_drawdown_gold
biggest_drawdown_dj
biggest_drawdown_btc

drawdowns_gold
drawdowns_dj
drawdowns_btc
```

## Conclusion

Gold recorded the smallest maximum drawdown (23.1%), falling from its peak in February 2026 to its trough in June 2026, indicating relatively stable price movements despite market uncertainty. The DJIA experienced a slightly larger drawdown (23.2%) between December 2019 and March 2020, corresponding to the global market crash triggered by the COVID-19 pandemic. Bitcoin suffered the largest drawdown (75.6%) from December 2017 to January 2019 during the cryptocurrency market downturn following the 2017 speculative bubble. Overall, the results support the hypothesis that volatility increases from Gold to DJIA and finally Bitcoin.

# Task 3: Visualising Singapore's Tourist Spending

## Load Data
```{r}
sg_tourism_data <- read.csv('singapore_tourism_receipts.csv')
```

## Visualisation Process
```{r}
p <- ggplot(sg_tourism_data, aes(x = value/1000, y = quarter, fill = item)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = c(
    'Sightseeing, Entertainment & Gaming'    = "#0B3D5C",
    'Shopping'       = "#2E86AB",
    'Other Components'          = "#E8CE5C",
    'Food & Beverage'  = "#E68A1E",
    'Accommodation'  = "#D64545"
  )) +
  geom_text(
    aes(label = ifelse(value/1000 >= 2.5, sprintf("%.3f", value/1000), "")),
    position = position_stack(vjust = 0.5),
    size = 2,
    color = "black"
  ) +
  guides(fill = guide_legend(nrow = 3, ncol = 2, byrow = FALSE)) +
  theme(legend.position = "top",
        legend.title = element_blank(),
        legend.background = element_blank(),
        legend.key.size = unit(0.25, "cm"),
        plot.background = element_rect(fill = "#FDF3DE", color = NA),
        panel.background = element_rect(fill = "white", color = NA),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()
        ) +
  labs(
    x = "Amount (Billion SGD)",
    y = "Quarter",
    title = "Singapore 2019-2024 Quarterly Tourism Receipts By Major Components",
    caption = 'Source: singstat.gov.sg'
  ) + 
  annotate("rect", xmin = 0, xmax = 10, ymin = 4, ymax = 14, alpha = .25) +
  annotate("text", x = 5, y = 10, label = "COVID-19 Pandemic") + 
  annotate("rect", xmin = 0, xmax = 30, ymin = 15, ymax = 25, alpha = .25) +
  annotate("text", x = 20, y = 21, label = "Tourism restrictions ease")
p
```

## Data Interpretation
During the COVID-19 pandemic, tourism numbers collapsed: 2021-Q1 shows the lowest total receipt (far below S\$1 billion, versus S\$8.026 billion for "Other Components" alone in 2019-Q4), reflecting Singapore's border closures.

Recovery was rapid once the pandemic subsided and borders reopened. 2023-Q4 saw a total receipt of S\$27.155 billion, nearly matching pre-pandemic levels (S\$29.192 billion in 2019-Q4).

Across all quarters, "Other Components" and "Sightseeing, Entertainment & Gaming" consistently form the largest shares, while "Food & Beverage" remains the smallest contributor (e.g. S\$3.192 billion in 2023-Q3 versus S$6.866 billion for "Other Components" that same quarter).


# Task 4: Reflection on the Use of GenAI

### What GenAI tools did we use, and how did they help?

Our team used ChatGPT, Claude and Little Bird AI throughout this assessment. We treated them as learning companions, coding assistants and debugging tools rather than relying on them to complete the assessment without understanding the work. Using more than one tool also allowed us to compare suggestions and choose the approach that best matched the task requirements.

For Task 1, ChatGPT helped us understand how to transform the EPL results from wide format into long format using pivot_longer(). It also guided us in separating the match scores into home and away goals, calculating points, goal difference and goals scored, and combining these statistics into the final league table. When designing the Home Advantage Index, GenAI suggested different possible formulas, but our team still had to decide which metric was most meaningful and justify our choice.

For Task 2, ChatGPT helped us clean the DJIA Excel file, convert the date column into R’s Date class, identify missing months and calculate monthly percentage changes. It also supported us in creating the combined violin and box plots, connecting monthly medians, and producing the faceted frequency polygons with rug plots and mean lines.

When comparing DJIA, Bitcoin and Gold, GenAI suggested using a logarithmic scale because the three assets had very different price ranges. It also helped us develop and verify the compute_drawdown() function using the provided test vector before applying it to the real data. During debugging, we shared screenshots and error messages with ChatGPT. For example, it helped us identify that tidyquant failed to load because Windows Application Control was blocking the TTR.dll dependency, rather than because our R code was incorrect.

For Task 3, we mainly used Claude to help us debug and recreate the tourism receipts visualisation. By sending Claude a screenshot of the reference chart, it could understand the overall layout, colours and formatting more effectively than if we had searched for each issue separately using a traditional search engine.

One specific example was the legend. We wanted to move it from the right side of the graph to the top, remove the legend title and adjust the text and key sizes so that it more closely resembled the reference visualisation. After we explained these requirements, Claude provided targeted suggestions for functions such as theme(), guides() and guide_legend().

Claude also helped us adjust labels, spacing and other visual elements. Little Bird AI was mainly used as a second source of suggestions when we wanted to compare possible approaches or confirm whether our explanation and visual design were reasonable.

### What did we find challenging?

The quality of the GenAI responses depended heavily on the quality of our prompts. When our questions were vague, the answers were often too general or did not solve the exact problem. We frequently had to provide screenshots, dataset structures, code drafts or full error messages before receiving a useful response.

We also found that GenAI tools could hallucinate or misunderstand certain R parameters. For Task 3, some suggested values for unit sizes, text alignment and scale names did not match the reference chart properly. As a result, we had to manually experiment with the code and combine parts of the AI-generated solution with lecture materials, official documentation and Stack Overflow.

Claude also went in an unnecessary direction when it initially suggested reshaping and rotating the graph into a wider format, even though the tourism dataset was already in a suitable long format. We decided not to follow that suggestion and instead adapted the useful parts of its code to fit our existing data structure.

Similarly, generated code could appear correct but still fail because of differences in package versions, file paths, column names or computer security settings. Therefore, we could not accept every answer without testing it in RStudio.

Another challen

> akash:
ge was that different tools sometimes suggested different solutions to the same problem. This meant that our team still needed to evaluate which method was logical, explainable and consistent with the lecture notes and assessment requirements.

### How did we use GenAI responsibly?

Our team followed this process:

1. Read the assessment instructions and relevant lecture notes.
2. Discuss the required output as a team.
3. Ask the GenAI tools for possible approaches or explanations.
4. Compare suggestions from different tools when necessary.
5. Run and test all code in RStudio.
6. Check whether the output matched the task requirements.
7. Modify the code, visualisations and explanations ourselves.
8. Verify the final calculations and interpretations before submission.

This was especially important for the visualisations. A plot could run successfully but still be misleading, cluttered or visually different from the required reference. We therefore checked the axis scales, legends, colours, labels, facet arrangement, annotations and overall readability ourselves.

### Overall reflection

Overall, ChatGPT, Claude and Little Bird AI improved our efficiency by reducing the time needed to search through forums, documentation and help threads. They also made debugging less frustrating because we could receive explanations that were specific to our code and screenshots.

However, this assessment also showed us that GenAI cannot replace human judgement. We still needed to understand the problem, test the code, reject unsuitable suggestions and decide whether the outputs were accurate and meaningful. The most effective approach was therefore a combination of GenAI assistance, comparison across tools, team discussion, technical testing and our own reasoning.
