---
title: "datavis-a2"
---

# 1. Setup libraries
```{r}
library(tidyverse)
library(readxl)
library(tidyquant)
```

# 2.1 DJIA Performance

## Import and clean the data (2%)

```{r}
raw_dj <- read_excel("dja-performance-report-monthly.xlsx", skip = 5)

monthly_dj <- raw_dj |>
  filter(`Index Name` == "Dow Jones Industrial Average") |>
  transmute(
    date = as.Date(`Effective Date`, format = "%m/%d/%Y"),
    close_value = as.numeric(`Close Value`)
  )

head(monthly_dj, 10)
```

## Identify the gap months (3%)

For every row, look at how many months until the
*next* row's date. If that's more than 1, there's a gap starting right there.
`lead()` peeks at the "next" value in a column (the opposite of `lag()`, which
looks backward).

```{r}
monthly_dj_check <- monthly_dj |>
  arrange(date) |>
  mutate(
    month = month(date),
    year = year(date),
    months_to_next_row = interval(date, lead(date)) %/% months(1)
  )

# The gap(s): any row where more than 1 month passes before the next entry
gap_start <- monthly_dj_check |>
  filter(months_to_next_row > 1)

gap_start
```

## Visualizing DJIA performance by month (10%)

- `rel_change_pct`: "how much did this month's close change vs last month's,
  in percent?" It's the classic `(new - old) / old * 100` formula. We use
  `lag()` to grab "last month's" value.
- `month_abb`: turns the numeric month (1–12) into "Jan", "Feb", ... and makes
  it an **ordered factor**. Without ordering, R would sort months
  alphabetically (Apr, Aug, Dec...) instead of calendar order — ordering fixes
  that so your x-axis reads Jan→Dec.

```{r}
monthly_dj <- monthly_dj |>
  arrange(date) |>
  mutate(
    rel_change_pct = (close_value / lag(close_value) - 1) * 100,
    month_abb = factor(month.abb[month(date)], levels = month.abb, ordered = TRUE)
  )

head(monthly_dj, 10)
```

### Visualization A: violin + box plot, with a line connecting medians

```{r viz-a}
monthly_dj_clean <- monthly_dj |> drop_na(rel_change_pct)

medians <- monthly_dj_clean |>
  group_by(month_abb) |>
  summarize(median_change = median(rel_change_pct), .groups = "drop")

ggplot(monthly_dj_clean, aes(x = month_abb, y = rel_change_pct)) +
  geom_violin(fill = "#a6cee3", color = "#1f78b4", trim = TRUE) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  geom_line(data = medians, aes(x = month_abb, y = median_change, group = 1),
            color = "firebrick", linewidth = 0.8) +
  geom_point(data = medians, aes(x = month_abb, y = median_change),
             color = "firebrick", size = 1.8) +
  labs(
    title = "DJIA Monthly Relative Change by Calendar Month",
    subtitle = "Violin shows full distribution shape; box shows quartiles; red line tracks the median",
    x = "Month", y = "Relative Change (%)"
  ) +
  coord_cartesian(ylim = quantile(monthly_dj_clean$rel_change_pct, c(0.01, 0.99))) +
  theme_minimal()
```

*(The `coord_cartesian(ylim = ...)` line just zooms the y-axis in on the
middle 98% of data so a few extreme historical crash months — like Oct
1929 — don't squash everything else flat. It doesn't delete any data, it just
changes the view, which is important since `coord_cartesian` — unlike
`ylim()` — never drops points from the stats calculations.)*

### Visualization B: faceted frequency polygons + rug plot

**Analogy:** a frequency polygon is just a histogram redrawn as a connected
line instead of bars — easier to compare shapes across many small panels. A
"rug plot" adds little tick marks along the bottom showing exactly where each
individual data point sits, like tick marks on a ruler.

`facet_wrap()` splits one big plot into a grid of small ones, one per group —
here, one panel per month.

```{r viz-b}
means <- monthly_dj_clean |>
  group_by(month_abb) |>
  summarize(mean_change = mean(rel_change_pct), .groups = "drop")

ggplot(monthly_dj_clean, aes(x = rel_change_pct)) +
  geom_freqpoly(bins = 30, color = "#1f78b4", linewidth = 0.7) +
  geom_rug(sides = "b", alpha = 0.3, color = "#1f78b4") +
  geom_vline(data = means, aes(xintercept = mean_change),
             linetype = "dashed", color = "firebrick") +
  facet_wrap(~ month_abb, nrow = 4, ncol = 3) +
  labs(
    title = "Distribution of DJIA Monthly Relative Change, by Month",
    subtitle = "Dashed red line = mean change for that month",
    x = "Relative Change (%)", y = "Count"
  ) +
  theme_minimal()
```

# 2.2 Comparing DJIA, BTC, and Gold Prices (25%)

## Import BTC and gold data (5%)

mport the BTC and gold price data since 2016 using the tq_transmute() function from the tidyquant package, and create `monthly_btc` and `monthly_gold` tibbles

```{r}
ticker_btc <- "BTC-USD"
ticker_gold <- "GC=F"
start_date <- "2016-01-01"
end_date <- "2026-06-30"

# BTC Data
monthly_btc <- tq_get(ticker_btc, from = start_date, to = end_date) %>%
  drop_na(adjusted) %>%
  tq_transmute(select = adjusted, mutate_fun = to.monthly, indexAt = "lastof")
# Gold Data
monthly_gold <- tq_get(ticker_gold, from = start_date, to = end_date) %>%
  drop_na(adjusted) %>%
  tq_transmute(select = adjusted, mutate_fun = to.monthly, indexAt = "lastof")

head(monthly_btc, 10)
head(monthly_gold, 10)
```

## Combined visualization (10%)

**Problem:** DJIA trades in tens of thousands, Gold in the
thousands, and BTC swings from under $1,000 to over $60,000 across this
period. Plot them raw on one chart and DJIA/BTC will flatten gold into an
invisible line at the bottom — like putting an elephant, a dog, and a mouse
on the same bathroom scale display.

**Solution: Indexing:** Rescale every series so it starts at 100. Now
we're not asking "what's the dollar price?" but "how much has this grown
since Jan 2016?" — a fair comparison. We also use a log y-axis, since BTC's
growth is multiplicative (10x, 100x) rather than additive, and log scales
make percentage moves visually comparable regardless of the starting size.

```{r}
dj_long <- monthly_dj |>
  filter(date >= as.Date(start_date)) |>
  transmute(month = floor_date(date, "month"), close_value, asset = "DJIA")

btc_long <- monthly_btc |>
  transmute(month = floor_date(as.Date(date), "month"), close_value = adjusted, asset = "BTC")

gold_long <- monthly_gold |>
  transmute(month = floor_date(as.Date(date), "month"), close_value = adjusted, asset = "Gold")

combined <- bind_rows(dj_long, btc_long, gold_long) |>
  drop_na(close_value)

# keep only the date range where ALL THREE assets have data
range_by_asset <- combined |>
  group_by(asset) |>
  summarize(min_month = min(month), max_month = max(month), .groups = "drop")

common_start <- max(range_by_asset$min_month)
common_end <- min(range_by_asset$max_month)

combined <- combined |>
  filter(month >= common_start, month <= common_end)

# index each series to 100 at the common start date
combined_indexed <- combined |>
  group_by(asset) |>
  arrange(month) |>
  mutate(index_100 = close_value / first(close_value) * 100) |>
  ungroup()

ggplot(combined_indexed, aes(x = month, y = index_100, color = asset)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  scale_color_manual(values = c(DJIA = "#1f78b4", BTC = "#e6550d", Gold = "#b15928")) +
  labs(
    title = "DJIA vs BTC vs Gold: Growth Since Jan 2016",
    subtitle = "Indexed to 100 at start, log scale so % moves are visually comparable",
    x = "Month", y = "Indexed Price (log scale, start = 100)", color = "Asset"
  ) +
  theme_minimal()
```

## Volatility analysis: largest monthly drawdown (10%)

`cummax()`: Walks through the price vector and records
"the highest price seen so far" at every point, then drawdown at each point = how far current price sits below
that high-water mark.

```{r}
compute_drawdown <- function(prices) {
  running_max <- cummax(prices)
  drawdown_pct <- (prices - running_max) / running_max * 100
  max_drawdown <- abs(min(drawdown_pct))
  max_drawdown
}

# Sanity check: peak of 120 -> trough of 90 is the biggest fall = -25%...
# wait, actually walk through it: 100 -> 60 (peak so far 100, DD -40%)
#                                  60 -> 120 (new peak, DD 0%)
#                                  120 -> 90 (peak 120, DD -25%)
# so the worst drawdown in this vector is -40%, matching the expected answer.
prices_test <- c(100, 60, 120, 90)
compute_drawdown(prices_test)  # should print 40
```

Now we extend the function slightly so it also tells us *when* the peak and
trough happened — useful for tying the numbers back to real events.

```{r drawdown-details}
compute_drawdown_details <- function(dates, prices) {
  running_max <- cummax(prices)
  drawdown_pct <- (prices - running_max) / running_max * 100
  trough_idx <- which.min(drawdown_pct)
  peak_idx <- max(which(prices[1:trough_idx] == running_max[trough_idx]))

  tibble(
    max_drawdown_pct = abs(drawdown_pct[trough_idx]),
    peak_date = dates[peak_idx],
    peak_price = prices[peak_idx],
    trough_date = dates[trough_idx],
    trough_price = prices[trough_idx]
  )
}

dj_since_2016 <- monthly_dj |>
  filter(date >= as.Date(start_date)) |>
  arrange(date)

drawdowns_dj <- compute_drawdown_details(dj_since_2016$date, dj_since_2016$close_value)
drawdowns_gold <- compute_drawdown_details(as.Date(monthly_gold$date), monthly_gold$adjusted)
drawdowns_btc <- compute_drawdown_details(as.Date(monthly_btc$date), monthly_btc$adjusted)

biggest_drawdown_dj <- drawdowns_dj$max_drawdown_pct
biggest_drawdown_gold <- drawdowns_gold$max_drawdown_pct
biggest_drawdown_btc <- drawdowns_btc$max_drawdown_pct

biggest_drawdown_gold
biggest_drawdown_dj
biggest_drawdown_btc

drawdowns_gold
drawdowns_dj
drawdowns_btc
```

**Writing your discussion paragraph:** once you run the chunk above, look at
the printed `peak_date` / `trough_date` for each asset and match them to
real events, e.g.:

- DJIA's worst 2016–2026 drawdown very likely centers on the **Feb–Mar 2020
  COVID-19 crash**.
- Gold's largest drawdown in this window is a plausible candidate for the
  **2022 Fed rate-hike cycle** (a stronger dollar and higher bond yields make
  non-yielding gold less attractive).
- BTC's largest drawdown is a plausible candidate for either the **Dec
  2017–Dec 2018 "crypto winter"** or the **Nov 2021–Nov 2022** collapse
  following the Terra/Luna and FTX failures.

Check your actual `peak_date`/`trough_date` output against these before
writing your final paragraph — don't state a date you haven't verified from
your own results. A ~100-word paragraph should: (1) state the three drawdown
percentages in increasing order, (2) name the peak/trough months per asset,
(3) link each to its real-world trigger, and (4) conclude whether the
ordering (Gold < DJIA < BTC in volatility) supports the hypothesis.
