---
title: "datavis-a2"
---

# 1. Setup libraries
```{r}
library(tidyverse)
library(readxl)
library(tidyquant)
```

# 2.1 DJIA Performance

## Import and clean the data (2%)

```{r}
raw_dj <- read_excel("dja-performance-report-monthly.xlsx", skip = 5)

monthly_dj <- raw_dj |>
  filter(`Index Name` == "Dow Jones Industrial Average") |>
  transmute(
    date = as.Date(`Effective Date`, format = "%m/%d/%Y"),
    close_value = as.numeric(`Close Value`)
  )

head(monthly_dj, 10)
```

## Identify the gap months (3%)

**Historical reason:** The New York Stock Exchange closed for over
four months (31 Jul–12 Dec 1914) at the outbreak of World War I, fearing
foreign investors would dump U.S. securities and drain gold reserves. No
trading meant no DJIA prices during that window, creating the gap in the
data.


```{r}
monthly_dj_check <- monthly_dj |>
  arrange(date) |>
  mutate(
    month = month(date),
    year = year(date),
    months_to_next_row = interval(date, lead(date)) %/% months(1)
  )

# The gap(s): any row where more than 1 month passes before the next entry
gap_start <- monthly_dj_check |>
  filter(months_to_next_row > 1)

gap_start
```

## Visualizing DJIA performance by month (10%)

- `rel_change_pct`: "how much did this month's close change vs last month's,
  in percent?"
- `month_abb`: turns the numeric month (1–12) into "Jan", "Feb", ... and makes
  it an **ordered factor**.

```{r}
monthly_dj <- monthly_dj |>
  arrange(date) |>
  mutate(
    rel_change_pct = (close_value / lag(close_value) - 1) * 100,
    month_abb = factor(month.abb[month(date)], levels = month.abb, ordered = TRUE)
  )

head(monthly_dj, 10)
```

### Visualization A: violin + box plot, with a line connecting medians

```{r viz-a}
monthly_dj_clean <- monthly_dj |> drop_na(rel_change_pct)

medians <- monthly_dj_clean |>
  group_by(month_abb) |>
  summarize(median_change = median(rel_change_pct), .groups = "drop")

ggplot(monthly_dj_clean, aes(x = month_abb, y = rel_change_pct)) +
  geom_violin(fill = "#a6cee3", color = "#1f78b4", trim = TRUE) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white") +
  geom_line(data = medians, aes(x = month_abb, y = median_change, group = 1),
            color = "firebrick", linewidth = 0.8) +
  geom_point(data = medians, aes(x = month_abb, y = median_change),
             color = "firebrick", size = 1.8) +
  labs(
    title = "DJIA Monthly Relative Change by Calendar Month",
    subtitle = "Violin shows full distribution shape; box shows quartiles; red line tracks the median",
    x = "Month", y = "Relative Change (%)"
  ) +
  coord_cartesian(ylim = quantile(monthly_dj_clean$rel_change_pct, c(0.01, 0.99))) +
  theme_minimal()
```

### Visualization B: faceted frequency polygons + rug plot

`facet_wrap()` splits one big plot into a grid of small ones, one per group —
here, one panel per month.

```{r viz-b}
means <- monthly_dj_clean |>
  group_by(month_abb) |>
  summarize(mean_change = mean(rel_change_pct), .groups = "drop")

ggplot(monthly_dj_clean, aes(x = rel_change_pct)) +
  geom_freqpoly(bins = 30, color = "#1f78b4", linewidth = 0.7) +
  geom_rug(sides = "b", alpha = 0.3, color = "#1f78b4") +
  geom_vline(data = means, aes(xintercept = mean_change),
             linetype = "dashed", color = "firebrick") +
  facet_wrap(~ month_abb, nrow = 4, ncol = 3) +
  labs(
    title = "Distribution of DJIA Monthly Relative Change, by Month",
    subtitle = "Dashed red line = mean change for that month",
    x = "Relative Change (%)", y = "Count"
  ) +
  theme_minimal()
```

# 2.2 Comparing DJIA, BTC, and Gold Prices (25%)

## Import BTC and gold data (5%)

Import the BTC and gold price data since 2016 using the tq_transmute() function from the tidyquant package, and create `monthly_btc` and `monthly_gold` tibbles

```{r}
ticker_btc <- "BTC-USD"
ticker_gold <- "GC=F"
start_date <- "2016-01-01"
end_date <- "2026-06-30"

# BTC Data
monthly_btc <- tq_get(ticker_btc, from = start_date, to = end_date) |>
  drop_na(adjusted) |>
  tq_transmute(select = adjusted, mutate_fun = to.monthly, indexAt = "lastof")
# Gold Data
monthly_gold <- tq_get(ticker_gold, from = start_date, to = end_date) |>
  drop_na(adjusted) |>
  tq_transmute(select = adjusted, mutate_fun = to.monthly, indexAt = "lastof")

head(monthly_btc, 10)
head(monthly_gold, 10)
```

## Combined visualization (10%)

**Problem:** DJIA trades in tens of thousands, Gold in the
thousands, and BTC swings from under $1,000 to over $60,000 across this
period. Plot them raw on one chart and DJIA/BTC will flatten gold into an
invisible line at the bottom — like putting an elephant, a dog, and a mouse
on the same bathroom scale display.

**Solution: Indexing:** Rescale every series so it starts at 100. Now
we're not asking "what's the dollar price?" but "how much has this grown
since Jan 2016?" — a fair comparison. We also use a log y-axis, since BTC's
growth is multiplicative (10x, 100x) rather than additive, and log scales
make percentage moves visually comparable regardless of the starting size.

```{r}
dj_long <- monthly_dj |>
  filter(date >= as.Date(start_date)) |>
  transmute(month = floor_date(date, "month"), close_value, asset = "DJIA")

btc_long <- monthly_btc |>
  transmute(month = floor_date(as.Date(date), "month"), close_value = adjusted, asset = "BTC")

gold_long <- monthly_gold |>
  transmute(month = floor_date(as.Date(date), "month"), close_value = adjusted, asset = "Gold")

combined <- bind_rows(dj_long, btc_long, gold_long) |>
  drop_na(close_value)

# keep only the date range where ALL THREE assets have data
range_by_asset <- combined |>
  group_by(asset) |>
  summarize(min_month = min(month), max_month = max(month), .groups = "drop")

common_start <- max(range_by_asset$min_month)
common_end <- min(range_by_asset$max_month)

combined <- combined |>
  filter(month >= common_start, month <= common_end)

# index each series to 100 at the common start date
combined_indexed <- combined |>
  group_by(asset) |>
  arrange(month) |>
  mutate(index_100 = close_value / first(close_value) * 100) |>
  ungroup()

ggplot(combined_indexed, aes(x = month, y = index_100, color = asset)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  scale_color_manual(values = c(DJIA = "#1f78b4", BTC = "#e6550d", Gold = "#b15928")) +
  labs(
    title = "DJIA vs BTC vs Gold: Growth Since Jan 2016",
    subtitle = "Indexed to 100 at start, log scale so % moves are visually comparable",
    x = "Month", y = "Indexed Price (log scale, start = 100)", color = "Asset"
  ) +
  theme_minimal()
```

## Volatility analysis: largest monthly drawdown (10%)

```{r}
compute_drawdown <- function(prices) {
  running_max <- cummax(prices)
  drawdown_pct <- (prices - running_max) / running_max * 100
  max_drawdown <- abs(min(drawdown_pct))
  max_drawdown
}

# Sanity check: peak of 120 -> trough of 90 is the biggest fall = -25%...
# wait, actually walk through it: 100 -> 60 (peak so far 100, DD -40%)
#                                  60 -> 120 (new peak, DD 0%)
#                                  120 -> 90 (peak 120, DD -25%)
# so the worst drawdown in this vector is -40%, matching the expected answer.
prices_test <- c(100, 60, 120, 90)
compute_drawdown(prices_test)  # should print 40
```

Now we extend the function slightly so it also tells us *when* the peak and
trough happened — useful for tying the numbers back to real events.

```{r drawdown-details}
compute_drawdown_details <- function(dates, prices) {
  running_max <- cummax(prices)
  drawdown_pct <- (prices - running_max) / running_max * 100
  trough_idx <- which.min(drawdown_pct)
  peak_idx <- max(which(prices[1:trough_idx] == running_max[trough_idx]))

  tibble(
    max_drawdown_pct = abs(drawdown_pct[trough_idx]),
    peak_date = dates[peak_idx],
    peak_price = prices[peak_idx],
    trough_date = dates[trough_idx],
    trough_price = prices[trough_idx]
  )
}

dj_since_2016 <- monthly_dj |>
  filter(date >= as.Date(start_date)) |>
  arrange(date)

drawdowns_dj <- compute_drawdown_details(dj_since_2016$date, dj_since_2016$close_value)
drawdowns_gold <- compute_drawdown_details(as.Date(monthly_gold$date), monthly_gold$adjusted)
drawdowns_btc <- compute_drawdown_details(as.Date(monthly_btc$date), monthly_btc$adjusted)

biggest_drawdown_dj <- drawdowns_dj$max_drawdown_pct
biggest_drawdown_gold <- drawdowns_gold$max_drawdown_pct
biggest_drawdown_btc <- drawdowns_btc$max_drawdown_pct

biggest_drawdown_gold
biggest_drawdown_dj
biggest_drawdown_btc

drawdowns_gold
drawdowns_dj
drawdowns_btc
```

## Conclusion

Gold recorded the smallest maximum drawdown (23.1%), falling from its peak in February 2026 to its trough in June 2026, indicating relatively stable price movements despite market uncertainty. The DJIA experienced a slightly larger drawdown (23.2%) between December 2019 and March 2020, corresponding to the global market crash triggered by the COVID-19 pandemic. Bitcoin suffered the largest drawdown (75.6%) from December 2017 to January 2019 during the cryptocurrency market downturn following the 2017 speculative bubble. Overall, the results support the hypothesis that volatility increases from Gold to DJIA and finally Bitcoin.


